---
title: "R Basics"
output: html_document
---

Objects are assigned values using the assignment operator. You can use "<-", "=", or "->" to assign values to objects, but "<-" is the preferred method. 

The basic data structure in R is a vector. Vectors are similar to arrays in other languages, but require uniform data. Vectors are created using the c() function.  

```{r}

numeric   <- c(1,2,3, 4)
numeric2  <- 1:4
char <- c("Tom", "Dick", "Harry")
logical   <- numeric > 2
not_allowed<- c(1:50, "Fabrice") #Numeric componetnts coerced to character. 

```

There are no scalars in R, if you assign an object only one number it is a vector of length one. 
```{r}
x <- 1
x
is.vector(x)

```

Matrixes are created by tacking together two vectors using cbind() or rbind()
```{r}
mtrx <- cbind(numeric, numeric2)
mtrx

```

Dataframes are basically matrixes with special properties, they are by far the most common menthod of working with data in R, and while a little slower than a matrix, are much more convenient.
```{r}

df <- data.frame(numeric, numeric2, c("a", "b", "c", "d"))
df
View(df)
mtcars<- mtcars #A built-in R dataset
mtcars
```




Elements of a vector or dataframe are accessed using square braces, a process known as subsetting. Indexes start at 1, not 0, which needs to be repeated because it causes so many problems INDEXES START AT 1 NOT 0!

```{r}
char
char[1:2]
char[c(1,1,1,1,3)]
char[3:1]
char[0]
char[0:3] #should probably throw an error, but doesn't!

#Data frames are accessed using [row number, column number]
mtcars[1, 5]
mtcars[1:3, 2:5]

#You can also subset by passing a logical or character vector to the [] operator. 

logic <- mtcars$cyl > 4
logic
mtcars[ logic, ]
mtcars[ , "cyl"]
mtcars[mtcars$cyl > 4, ]


```

Since accessing variables is so common, $ is used as shorthand for a variable name. mtcars$cyl, is the same as mtcars[, "cyl"].

All of your favourite mathmatical and logical operators exist in R:

```{r}
x<- 1:5
x
x > 2
x < 2
x >= 2
x == 2 #two '=', not three. 
x != 2
!(x > 2)

2 %in% x #tests if left hand side is present in right hand side. 

TRUE & FALSE
TRUE | FALSE

1 + 1
1 - 1
2 * 3
4 / 2
4 ^ 2
5 %% 2 #modulus

```

For and While loops are executed in R, so are very slow. Nine times out of ten there is a vectorized solution to the problem you are trying to solve with a loop. Using that solution will speed up your code and also make it more readable. It's also easy to replace R loops with C++ loops using the RCpp package. 

```{r}

for(i in 1:5){
  print(i)
}

#For loops iterates through each element of the vector. To change how the loop iterates,
#change the vector. 

for(i in c("a", "b", "c", "methuzelah")){
  print(i)
}

for(i in seq(from = 0, to = 10, by = 2)){
  print(i)
}


j<- 1
while( j < 5){
  print(j)
  j<- j+1
}

```

If/Else statements 

```{r}

for( i in 1:10){
  if(i < 2){
    print("Plato")
  }else if(i < 6){
    print("Hericlitus")
  }else
    print("Theracles")
}

x<- 1:10
ifelse(x > 5, "big", "small")

```

The main flow control operators in R are the break and next commands

```{r}

for(i in 1:10){
  if(i == 3){
    next
  }else if(i == 7){
    break
  }else
    print(i)
}
```

R is a mostly functional programming language, and behaves better if you treat it like one. Function definition looks a lot like Javascript. 

```{r}

doubler <- function(x){
  x * 2
}

doubler(4)
doubler(1:10)

#use the equals sign to set default value of funcitons

adder <- function(x, y = 23, text = "Nice Addition"){
  print(x + y)
  print(text)
}

adder(2)
adder(2, y = 5, text = "Huzzah!")
adder(2, 3, "Huzzah!")

#closures are functions that return functions

power <- function(x){
  function(y){
    y ^ x
  }
}

square <- power(2)
cube <- power(3)
square(1:6)
cube(1:5)
```


R has many wyas to read and write data. 

```{r}

write.csv(mtcars, "mtcars.csv")
df<- read.csv("mtcars.csv")

```








